---
title: "Assignment 1 Spatial Epidemiology"
author: "Ferrara Lorenzo, Lucchini Marco"
date: "25-10-2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Elevation of a small islet in Delta d’Ebre
---


```{r include=FALSE}
knitr::opts_chunk$set( echo=F )
knitr::opts_chunk$set(out.width = '70%')
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(tidy =TRUE)
knitr::opts_chunk$set(message = FALSE)
```

# Description

The data were collected during a study of the settlement pattern of common terns on a small islet in the Delta d’Ebre (Hernandez and Ruiz, range3), particularly in the mouths of the Ebre river. The islet was inspected at two-day intervals throughout the range0 breeding season. The data include the location of each nest, its elevation above sea level, and elevations at a number of additional points (points without nest) on the islet. In the file called elevationsIslet.txt, contains the information of the coordinates and elevation above sea, and in file, called poly84.txt contains the coordinates of the borders of the islet. The aim is to predict the elevation above sea level along the small islet using a kriging interpolation. 


```{r eval=FALSE, include=FALSE}
# load("Data.RData")
```

```{r, include=FALSE}
# setwd("~/Library/CloudStorage/OneDrive-PolitecnicodiMilano/Polimi/5.1 Bayesian methods - Oriol/GEOSTAT/Geostat/Task 1/GIT/Epidemiology_Assignment_1/")
poly84 <- read.delim("poly84.txt")
names(poly84)= c("x", "y", "data")
# head(poly84)
dataset <- read.delim("elevationsIslet.txt")
# head(elevationsIslet)
```

## 1) Explore the requirement of stationary mean of the process. In case this requirement is not met, detrend the data to ensure that the process is stationary in mean. Discuss the results and show the plot of the results

Firstly, we plot the locations to see the spatial distribution of the data

```{r include=FALSE}
library(geoR)
library(gstat)
# library(lattice)
library(sp)
```

```{r}
coordinates(dataset) <- c('x','y')
#bubble(dataset,'data', col=dataset$data, key.space='bottom')

rbPal <- colorRampPalette(c('green','black'))
col_scale <- rbPal(10)[as.numeric(cut(dataset$data,breaks = 10))]
rbPal_neg <- colorRampPalette(c('Red','pink'))
col_scale[dataset$data<0] = rbPal_neg(10)[as.numeric(cut(dataset$data[dataset$data<0],breaks = 10))]
plot(dataset, pch = 20, col=col_scale,  ylim = c(-90, 89), main = "Elevation of the islet")
lines(poly84, lwd=2)
legend('bottomright', pch = 16, c("-9.5", "0", "7", "18", "39.6"), col=c("#FF0000", "#FF959D", "#00E200" ,"#005500", "#001C00"))
```


We also look at the distribution in relation to the x-coordinates (E-W) and y-coordinates(N-S).

```{r}
geodataset <- as.geodata(dataset)
#plot(geodataset, borders=poly84[c("x","y")], trend="2nd")
par(mfrow=c(1,2))
with(geodataset , plot(coords[, 1], data, xlab = "x",ylab = "data", pch = 20, cex = 0.7))
#add lowess smoother
lines(lowess(geodataset$data ~ geodataset$coords[, 1]))

with(geodataset , plot(coords[, 2], data, xlab = "y",ylab = "data", pch = 20, cex = 0.7))
#add lowess smoother
lines(lowess(geodataset$data ~ geodataset$coords[, 2]))
```

The plots show a concentration of high values in the extreme east and west of the islet and we can observe a greater density in the north.

The process doesn't seem stationary, indeed there is a clear quadratic trend along the x direction. In addition we try using the y direction as regressor, to find a possible linear or quadratic trend.

```{r  }
lm <- lm(data ~ x + y + I(x^2) + I(y^2), data = dataset)
lm <- lm(data ~ x + y + I(x^2), data = dataset)
summary(lm)
```

The linear term in y doesn't seem significant, so we remove it.

```{r }
lm2 <- lm(data ~ x + I(x^2) + I(y^2), data = dataset)
summary(lm2)
```

Now we have obtained a model in which all regressors seem to be significant. We save the residuals of our linear model and look at the de-trended data:

```{r  }
df <- read.delim("elevationsIslet.txt")
residuals <- round(residuals(lm2), digits = 3)
dataset.new <- cbind(df, residuals)
# head(dataset.new)


geodataset.residuals <- as.geodata(dataset.new, data.col = 4)

# plot(geodataset.residuals, borders=poly84[c("x","y")])  # trend="2nd"
```


```{r include = FALSE}
rbPal <- colorRampPalette(c('red', 'green', 'blue', 'black'))
col_scale <- rbPal(4)[as.numeric(cut(geodataset.residuals$data,breaks = 4))]
plot(dataset, pch = 20, col=col_scale, xlim = c(-2.1, 124), ylim = c(-90, 89), main = "Detrended elevation of the islet")
lines(poly84, lwd=2)
legend('topright', pch = 16, c("1st", "2nd", "3rd", "4th"), col=c('red', 'green', 'blue', 'black'), title="Quantile")
```

```{r eval=FALSE, include=FALSE}
plot(geodataset.residuals)
```

What we now obtain is a more homogeneous distribution of the data around the value 0.

<!-- NON E' COERENTE DAL QUELLO OTTENUTO CON PLOT -->


```{r}
par(mfrow=c(1,2))
with(geodataset.residuals , plot(coords[, 1], data, xlab = "x",ylab = "data", pch = 20, cex = 0.7))
#add lowess smoother
lines(lowess(geodataset.residuals$data ~ geodataset.residuals$coords[, 1]))

with(geodataset.residuals , plot(coords[, 2], data, xlab = "y",ylab = "data", pch = 20, cex = 0.7))
#add lowess smoother
lines(lowess(geodataset.residuals$data ~ geodataset.residuals$coords[, 2]))
```

We don't notice any particular trend so this new dataset seems stationary!

<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

\newpage

## 2) Explore the spatial dependence of the elevation variable using the variogram cloud and bins and the empirical variogram. Discuss the results and plot them

We try to analyse the spatial dependence through a Variogram Cloud, using the Robust Estimator (since the Classical one might consider outliers some values which are not necessarily outliers)

```{r results='hide'}
vario.cloud.robust <- variog(geodata = geodataset.residuals, option = "cloud", estimator.type = "modulus")
plot(vario.cloud.robust, main = "CLOUD", cex.main = 1, cex.lab = 1)
```

The variability at small distances appears a bit greater than that for larger distances. 
=> We reduce the density of the plot by reducing the maximum distance over which the variances are calculated. 

```{r results='hide'}
dist.max.data  <- max(vario.cloud.robust$u)/2

vario.cloud.robust <- variog(geodata = geodataset.residuals,option = "cloud", estimator.type = "modulus", max.dist = dist.max.data)

plot(vario.cloud.robust, main = "CLOUD", cex.main = 1, cex.lab = 1)
```

We also have a look at the boxplots:

```{r results='hide' }

vario.bc.robust <- variog(geodata = geodataset.residuals, option = "bin",bin.cloud = TRUE, pairs.min = 30, max.dist = dist.max.data, estimator.type = "modulus")

plot(vario.bc.robust, bin.cloud = TRUE, cex.lab = 1, main = "\nBINNED BOXPLOTS",cex.main = 1)
```

```{r results='hide' }
vario.bc.robust$ind.bin
```

=> All the bins have at least $pairs.min=30$ observations each, indeed they have:

```{r}
vario.bc.robust$n
```

Therefore the Empirical Variogram is:

```{r results='hide' }
vario.b.robust <- variog(geodata = geodataset.residuals, option = "bin",pairs.min = 30, max.dist = dist.max.data, estimator.type = "modulus",uvec = seq(1, dist.max.data, l = 13))

plot(vario.b.robust, main = "EMPIRICAL VARIOGRAM (Robust)\nBINNED",cex.main = 1, cex.lab = 1, cex = 1, pch = 16)
```

From what we can observe the variogram of the residuals the values increases until certain distance, and then they keep constant. That's ok since it's the behaviour that we expect from a stationary variogram.

\newpage

## 3) Check the hypothesis of the spatial independence

To check the hypothesis of the spatial independence we use a Monte Carlo approach 

```{r  }
set.seed(5000) #fix the seed
indep.env <- variog.mc.env(geodataset.residuals, obj.variog = vario.b.robust,nsim = 1500)
plot(vario.b.robust, envelope = indep.env, main = "CONFIDENCE BANDS FOR INDEPENDENT MODEL",lwd = 2, pch = 16)
```

All the values of the empirical variogram are inside the envelope, therefore the process has no spatial dependence.

\newpage

## 4) Check the isotropy property of the process. Comment the results, it’s not necessary to overcome the anisotropy.

To check the anisotropy we need to compute the directional variogram in the 4 main directions: 0º, 45º,90º and 135º. 

<!-- classical -->

```{r eval=FALSE, include=FALSE, results='hide'}
variod <- variog4(geodata = geodataset.residuals, option = "bin",pairs.min = 30, max.dist = dist.max.data, estimator.type = "classical",uvec = seq(1, dist.max.data, l = 13))

plot(variod, lyt = 2, legend = FALSE)
#plot the directional variogram
legend(x = "bottomright", inset = 0.01, lty = c(1, 2, 3, 4),col = c("black", "red", "green", "blue"), legend = c("0º","45º", "90º", "135º"), cex = 0.5)
```

<!-- oppure modulus? -->

```{r  results='hide'}
variod <- variog4(geodata = geodataset.residuals, option = "bin",pairs.min = 30, max.dist = dist.max.data, estimator.type = "modulus",uvec = seq(1, dist.max.data, l = 13))

plot(variod, lyt = 2, legend = FALSE)
#plot the directional variogram
legend(x = "bottomright", inset = 0.01, lty = c(1, 2, 3, 4),col = c("black", "red", "green", "blue"), legend = c("0º","45º", "90º", "135º"), cex = 0.5)
```

The directional variograms don't seem to be perfectly overlapping: they might have the a different sill, in particular the variogram seems to have a higher value along the 90° direction => Geometrical anistropy

Let's also analyse the range observing the Rose Diagram:

```{r  results='hide'}
source("RoseDiagram.R")

NumCases = length(variod$`0`)

# crit.val~sill
# max.dist = massimo valore di dstanza
rose.diagram(data.var = geodataset.residuals$data, data.cds = geodataset.residuals$coord,
max.dist = 90, numcases = NumCases, numdirec = 4, poly.tnd = "cte",
crit.val = 120)

```


We notice that the Rose diagram is not perfectly circular: it is slightly elliptical, with major range in the 45° and 90° direction => Zonal anistropy.

=> In conclusion we have a combined anisotropy, so we can't make the assumption of isotropic process, which is necessary for the correct use of the kriging techniques, since the theoretical variograms used for krigring are based on isotropic models.

\newpage
## 5) Propose four theoretical variograms and estimate the parameters via restricted maximum likelihood or weighed least square. Select the two variograms which best fit the data. Explain the parameters of the chosen variogram (sill, nugget, range and kappa).

<!-- The κ parameter controls the smoothness of the process. -->
<!-- The higher is κ, the smoother the spatial process. -->
<!-- Special cases κ = 1/2 exponential model and κ → ∞ Gaussian mode -->

```{r results='hide'}
variogramma <- variog(geodata = geodataset.residuals, estimator.type = "modulus")
```

```{r eval=FALSE, include=FALSE}
windows(); eyefit(variogramma, silent = FALSE)
```


We'll use a Maximum Likelihood approach

```{r include=FALSE}
lk.exponential <- likfit(geodataset.residuals, cov.model = "exponential", ini = c(91, 25), lik.method = "REML")

lk.gaussian <- likfit(geodataset.residuals, cov.model = "gaussian", ini = c(101, 11), lik.method = "REML")

lk.spherical <- likfit(geodataset.residuals, cov.model = "spherical", ini = c(91, 22), lik.method = "REML")

lk.matern <- likfit(geodataset.residuals, cov.model = "matern", ini = c(88, 19), fix.nugget = F, fix.kappa = F, kappa = 0.14, lik.method = "REML")
```

### Exponential
```{r  }
lk.exponential
```

### Gaussian
```{r  }
lk.gaussian
```

### Spherical
```{r  }
lk.spherical
```


### Matern
```{r  }
lk.matern
```

```{r  }

dist.max.data=149

plot(variogramma, main = "PARAMETRIC VARIOGRAMS", cex.main = 1, pch = 16) #empirical variogram
lines(lk.exponential, lwd = 2, col = "red", max.dist = dist.max.data) # add exponential
lines(lk.gaussian, lwd = 2, col = "blue", max.dist = dist.max.data) # add guassian
lines(lk.spherical, lwd = 2, col = "green3", max.dist = dist.max.data) # add spherical
lines(lk.matern, lwd = 2, col = "yellow", max.dist = dist.max.data) # add matern
legend(x = "bottomright", inset = 0.01, lty = c(1, 1), col = c("red",
"blue", "green3", "yellow"), legend = c("Exponential", "Gaussian",
"Spherical", "Matern"), cex = 1)
```

```{r  , include=FALSE}
env.exponential <- variog.model.env(geodataset.residuals, obj.variog = variogramma, model.pars = lk.exponential)
env.gaussian <- variog.model.env(geodataset.residuals, obj.variog = variogramma, model.pars = lk.gaussian)
env.spherical <- variog.model.env(geodataset.residuals, obj.variog = variogramma, model.pars = lk.spherical)
env.matern <- variog.model.env(geodataset.residuals, obj.variog = variogramma, model.pars = lk.matern)
```

```{r  }
par(mfrow = c(1, 4))
plot(variogramma, main = "Exponential", lwd = 2, pch = 16, envelope = env.exponential)
plot(variogramma, main = "Gaussian", lwd = 2, pch = 16, envelope = env.gaussian, cex.main = 1)
plot(variogramma, main = "Spherical", lwd = 2, pch = 16, envelope = env.spherical, cex.main = 1)
plot(variogramma, main = "Matern", lwd = 2, pch = 16, envelope = env.matern, cex.main = 1)
```

All the simulated variograms contain the empirical variogram so we don't have evidence to exclude any model.

Let's compare the different models
```{r  }

temp.lk = data.frame( model = c("exponential", 
                      "gaussian", 
                      "spherical", 
                      "matern"), 
            loglikelihood = c(lk.exponential$loglik, 
                               lk.gaussian$loglik, 
                               lk.spherical$loglik,
                               lk.matern$loglik
                               )
)

temp.lk[ order(temp.lk$loglikelihood, decreasing = T),]
```

The two fitted models with the highest loglikelihood are the Matern model and the Gaussian model. So we'll use these two to perform the kriging prediction step.

\newpage
## 6) Predict the elevations along all the area of study using the two variogram selected in point 4. Discuss the type of kriging chosen:
### a. Compare both kriging predictions using cross-validation, and propose the best model.
### b. Show the predictions and their standard errors. 

First generate a grid where we'll perform our kriging predictions

```{r  }
rnx <- range(geodataset$coords[,1]) #Minimum and maximum of the latitude
rny <- range(geodataset$coords[,2]) #Minimum and maximum of the longitude
newx.grid <- seq(rnx[1],rnx[2],l=51) #Create a sequence from minimum to
#maximum of the latitude of 51 points
newy.grid <- seq(rny[1],rny[2],l=51) #Create a sequence from minimum to
#maximum of the longitude of 51 points
dsgr.grid <- expand.grid(newx=newx.grid, newy=newy.grid) #expand, to create a grid

points(geodataset, ylim=range(poly84$y))
lines(poly84[,1:2], type="l")
points(dsgr.grid, pch = 19, col = 4, cex = 0.25)
```

#### Gaussian

Let's try using the Gaussian model to do our prediction step:

```{r  }
kc.gaussian<-krige.conv(geodataset,locations= dsgr.grid,
                        krige = krige.control(type.krige="OK",obj.m = lk.gaussian, trend.l="2nd",trend.d= "2nd"))

image(kc.gaussian,xlim=range(poly84$x), ylim=range(poly84$y), borders = poly84)
```

```{r  }
contour(kc.gaussian,filled= TRUE,
coords.data=geodataset$coords,col=terrain.colors(31),xlim=range(poly84$x),ylim=range(poly84$y),borders = poly84)
```

```{r  }
contour(kc.gaussian,filled= TRUE,val=sqrt(kc.gaussian$krige.var),
coords.data=geodataset$coords,xlim=range(poly84$x),ylim=range(poly84$y),
col = gray(seq(1, 0.1, l = 30)), main="Plot of Standard Deviation",borders = poly84)
```
 
```{r  }
xv.gaussian <- xvalid(geodataset, model =lk.gaussian, reestimate = F)

names(xv.gaussian)
```

```{r  }
VC1.gaussian <- abs(mean(xv.gaussian$error/sqrt(xv.gaussian$krige.var)))
VC2.gaussian <- sqrt(mean((xv.gaussian$error/sqrt(xv.gaussian$krige.var))^2))
VC3.gaussian <- sqrt(mean(xv.gaussian$error^2))

data.frame( index = c("VC1", "VC2", "VC3"), value = c(VC1.gaussian, VC2.gaussian, VC3.gaussian) )

```


<!-- ############################ -->

#### Matern 
```{r  }
kc.matern<-krige.conv(geodataset,locations= dsgr.grid,
krige = krige.control(type.krige="OK",obj.m = lk.matern,  trend.l="2nd", trend.d= "2nd"))
image(kc.matern, xlim=range(poly84$x),ylim=range(poly84$y), borders = poly84)
```
```{r  }
contour(kc.matern,filled= TRUE,
coords.data=geodataset$coords,xlim=range(poly84$x),ylim=range(poly84$y), col=terrain.colors(30), borders = poly84)
```

```{r  }
contour(kc.matern,filled= TRUE,val=sqrt(kc.matern$krige.var),
coords.data=geodataset$coords,xlim=range(poly84$x),ylim=range(poly84$y), col = gray(seq(1, 0.1, l = 30)), borders = poly84[,1:3])
```

```{r  }
xv.matern <- xvalid(geodataset, model =lk.matern, reestimate = F)
```

```{r  }
VC1.matern <- abs(mean(xv.matern$error/sqrt(xv.matern$krige.var)))
VC2.matern <- sqrt(mean((xv.matern$error/sqrt(xv.matern$krige.var))^2))
VC3.matern <- sqrt(mean(xv.matern$error^2))

data.frame( index = c("VC1", "VC2", "VC3"), value = c(VC1.matern, VC2.matern, VC3.matern) )
```








```{r  }
save.image("Data.RData")
```




